---
title: "Runtime Contracts for R Functions"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Runtime Contracts for R Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dev = "svglite",
  fig.ext = "svg",
  error = TRUE
)
```

```{r setup}
library(restrictR)
```

## Overview

`restrictR` lets you define reusable input contracts from small building blocks
using the base pipe `|>`. A contract is defined once and called like a function
to validate data at runtime. Validators are immutable: each `|>` returns a new
validator, so you can safely branch from a shared base without side effects.

| Section | What you'll learn |
|---------|-------------------|
| [Reusable schemas](#reusable-schemas) | Define and reuse data.frame contracts |
| [Dependent validation](#dependent-validation) | Constraints that reference other arguments |
| [Enum arguments](#enum-arguments) | Restrict string arguments to a fixed set |
| [Data frame with mixed constraints](#data-frame-with-mixed-constraints) | Columns + enums + ranges in one contract |
| [Custom steps](#custom-steps) | Domain-specific invariants |
| [Self-documentation](#self-documentation) | Print, `as_contract_text()`, `as_contract_block()` |
| [Using contracts in packages](#using-contracts-in-packages) | The recommended pattern for R packages |

## Reusable Schemas

The most common use case: validating a `newdata` argument in a predict-like
function. Instead of scattering `if`/`stop()` blocks, define the contract once:

```{r}
require_newdata <- restrict("newdata") |>
  require_df() |>
  require_has_cols(c("x1", "x2")) |>
  require_col_numeric("x1", no_na = TRUE, finite = TRUE) |>
  require_col_numeric("x2", no_na = TRUE, finite = TRUE) |>
  require_nrow_min(1L)
```

The result is a callable function. Valid input passes silently:

```{r}
good <- data.frame(x1 = c(1, 2, 3), x2 = c(4, 5, 6))
require_newdata(good)
```

Invalid input produces a structured error with the exact path and position:

```{r}
require_newdata(42)
```

```{r}
require_newdata(data.frame(x1 = c(1, NA), x2 = c(3, 4)))
```

```{r}
require_newdata(data.frame(x1 = c(1, 2), x2 = c("a", "b")))
```

Every error follows the same format: `path: message`, optionally followed by
`Found:` and `At:` lines. This makes errors instantly recognizable and
grep-friendly.

## Dependent Validation

Some contracts depend on context. A prediction vector must have the same length
as the rows in `newdata`:

```{r}
require_pred <- restrict("pred") |>
  require_numeric(no_na = TRUE, finite = TRUE) |>
  require_length_matches(~ nrow(newdata))
```

The formula `~ nrow(newdata)` declares a dependency on `newdata`. Pass it
explicitly when calling the validator:

```{r}
newdata <- data.frame(x1 = 1:5, x2 = 6:10)
require_pred(c(0.1, 0.2, 0.3, 0.4, 0.5), newdata = newdata)
```

Mismatched lengths produce a precise diagnostic:

```{r}
require_pred(c(0.1, 0.2, 0.3), newdata = newdata)
```

Missing context is caught **before any checks run**:

```{r}
require_pred(c(0.1, 0.2, 0.3))
```

Context can also be passed as a named list via `.ctx`:

```{r}
require_pred(1:5, .ctx = list(newdata = newdata))
```

## Enum Arguments

For string arguments that must be one of a fixed set:

```{r}
require_method <- restrict("method") |>
  require_character(no_na = TRUE) |>
  require_length(1L) |>
  require_one_of(c("euclidean", "manhattan", "cosine"))
```

```{r}
require_method("euclidean")
```

```{r}
require_method("chebyshev")
```

## Data Frame with Mixed Constraints

Contracts work well for functions that accept a data frame with typed columns,
value ranges, and categorical fields in one go:

```{r}
require_survey <- restrict("survey") |>
  require_df() |>
  require_has_cols(c("age", "income", "status")) |>
  require_col_numeric("age", no_na = TRUE) |>
  require_col_between("age", lower = 0, upper = 150) |>
  require_col_numeric("income", no_na = TRUE, finite = TRUE) |>
  require_col_one_of("status", c("active", "inactive", "pending"))
```

```{r}
good_survey <- data.frame(
  age = c(25, 40, 33),
  income = c(35000, 60000, 45000),
  status = c("active", "inactive", "active")
)
require_survey(good_survey)
```

```{r}
bad_survey <- data.frame(
  age = c(25, -5, 200),
  income = c(35000, 60000, 45000),
  status = c("active", "inactive", "active")
)
require_survey(bad_survey)
```

## Custom Steps

For domain-specific invariants that don't belong in the built-in set, use
`require_custom()`. The step function receives `(value, name, ctx)` and should
call `fail()` on failure to produce the same structured errors as built-in
steps:

```{r}
require_weights <- restrict("weights") |>
  require_numeric(no_na = TRUE) |>
  require_between(lower = 0, upper = 1) |>
  require_custom(
    label = "must sum to 1",
    fn = function(value, name, ctx) {
      if (abs(sum(value) - 1) > 1e-8) {
        fail(name, "must sum to 1",
             found = sprintf("sum = %g", sum(value)))
      }
    }
  )
```

```{r}
require_weights(c(0.5, 0.3, 0.2))
```

```{r}
require_weights(c(0.5, 0.5, 0.5))
```

Custom steps can also declare dependencies:

```{r}
require_probs <- restrict("probs") |>
  require_numeric(no_na = TRUE) |>
  require_custom(
    label = "length must match number of classes",
    deps = "n_classes",
    fn = function(value, name, ctx) {
      if (length(value) != ctx$n_classes) {
        fail(name, sprintf("expected %d probabilities", ctx$n_classes),
             found = sprintf("length %d", length(value)))
      }
    }
  )

require_probs(c(0.3, 0.7), n_classes = 2L)
```

## Self-Documentation

Print a validator to see its full contract:

```{r}
require_newdata
```

Use `as_contract_text()` to generate a one-line summary for roxygen `@param`:

```{r}
as_contract_text(require_newdata)
```

Use `as_contract_block()` for multi-line output suitable for `@details`:

```{r}
cat(as_contract_block(require_newdata))
```

## Using Contracts in Packages

The recommended pattern: define contracts near the top of the file that uses
them, or in a dedicated `R/contracts.R` if several files share the same
validators. Call them at the top of exported functions.

```r
# R/contracts.R
require_newdata <- restrict("newdata") |>
  require_df() |>
  require_has_cols(c("x1", "x2")) |>
  require_col_numeric("x1", no_na = TRUE, finite = TRUE) |>
  require_col_numeric("x2", no_na = TRUE, finite = TRUE)

require_pred <- restrict("pred") |>
  require_numeric(no_na = TRUE, finite = TRUE) |>
  require_length_matches(~ nrow(newdata))
```

```r
# R/predict.R

#' Predict from a fitted model
#'
#' @param newdata `r as_contract_text(require_newdata)`
#' @param ... additional arguments passed to the underlying model.
#'
#' @export
my_predict <- function(object, newdata, ...) {
  require_newdata(newdata)
  pred <- do_prediction(object, newdata)
  require_pred(pred, newdata = newdata)
  pred
}
```

Contracts compose naturally with the pipe and branch safely (each `|>` creates a
new validator):

```{r}
base <- restrict("x") |> require_numeric()
v1 <- base |> require_length(1L)
v2 <- base |> require_between(lower = 0)

# base is unchanged
length(environment(base)$steps)
length(environment(v1)$steps)
length(environment(v2)$steps)
```

```{r}
sessionInfo()
```
